Lessons learned and paths forward for mass dog vaccination in Madagascar: a case study of pilot canine rabies vaccination campaigns in Moramanga District

### 2021-02-03 manuscript.Rmd
adding intermediate results to out

---
title: Lessons learned from pilot canine rabies vaccination campaigns in Moramanga
  District, Madagascar
author: "Malavika Rajeev"
date: "11/19/2020"
output:
  word_document: default
  html_document: default
bibliography: format/references.bib
---

@@

Canine rabies results in an estimated 60,000 human deaths per year globally. These deaths are entirely preventable: prompt post-exposure prophylaxis (PEP) of humans is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination. Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in domestic dog populations. However, achieving this coverage target in low and middle income countries where the virus can be challenging due to economic, sociocultural, and political reasons.

@@

Here, we summarize lessons learned through implementation of pilot vaccination programs in the Moramanga District of Madagascar, where previous work has shown high incidence of dog rabies cases and human exposures. In 2018 and 2019, we deployed two different vaccination strategies. In 2018, we led a larger scale volunteer and NGO led pilot vaccination campaign in two Communes (sub-district level) using a central point strategy. In 2019/2020, we provided vaccinations and a per-vaccine fee to the District Veterinarian (DV) to vaccinate animals as part of routine services. We compare monetary costs and coverage estimates between the campaigns. Using demographic data collected during the campaigns and a vaccination model, we explore different vaccination strategies based on what we learned during implementation.

@@

*Willingness to pay*.

*Person days*.

*Vax per point*.

@@

### Summary of 2018 and 2019 campaigns
```{r}
species_2018 <- table(animals_2018$Species)
species_2019 <- table(tolower(animals_2019$`Dog or Cat`))

@@

*Table 1 should describe dogs!! (Take out cat parts & keep dogs only, & person days!)*

@@

@@

In Andasibe Ville, we were only able to did this in one of our campaign locations, due to the sparsely distributed households and roads throughout the area, as such coverage in this

### Comparing campaign costs and willingness to pay
Costs (different categories)
Person days + dogs vax per person day (number of vaccination points)
Willingness to pay

### Dog demography and simulating vaccination strategies

For dogs, the sex ratio was male-biased (55%); while for cats it was female-biased (68% female). For dogs, the sex ratio was male-biased (65%), while for cats it was female-biased (66% female).

```{r this is a code chunk!}

@@

More community based delivery (Ahmed's paper, Pivot, accessibility, etc.). Possibly combining different campaign strategies could work per models Turnover is tricky & waning doesn't contribute too much. Thermotolerant & 3 yrs even without booster. (although most dogs are revaccinated and we do this in the model). Volunteer based/NGO based. Push for govt investment and viewing as a public good. Modeling after childhood vax = routine through local offices of puppies + supplemental activities by trained vets annually to catch unvax (feral, oral). The veterinary workforce (relieving pressures on them, challenges with private-public roles).

@@

@@

A note abt Annie. Others who helped

>>>>

---
title: "Lessons learned and paths forward for mass dog vaccination in Madagascar: a case study of pilot canine rabies vaccination campaigns in Moramanga District"
author: "Malavika Rajeev"
date: "11/19/2020"
output:
  word_document: default
  html_document: default
bibliography: format/references.bib
---

@@

Canine rabies results in an estimated 60,000 human deaths per year globally. These deaths are entirely preventable: prompt post-exposure prophylaxis (PEP) of humans is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination. Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in domestic dog populations. However, achieving this coverage target in low and middle income countries where the virus ia endemic can be challenging due to economic, ecological sociocultural, and political reasons.

@@

Here, we summarize lessons learned through implementation of pilot vaccination programs in the Moramanga District of Madagascar, where previous work has shown high incidence of dog rabies cases and human exposures. In 2018 and 2019, we deployed two different vaccination strategies. In 2018, we led a larger scale volunteer-led pilot vaccination campaign in two Communes (sub-district level) using a central point strategy. In 2019/2020, we provided vaccinations and a per-vaccine fee to the District Veterinarian (DV) to vaccinate animals as part of routine services. We compare monetary costs and coverage estimates between the campaigns. Using demographic data collected during the campaigns and a vaccination model, we explore different vaccination strategies based on what we learned during implementation.

@@

*Willingness to pay*. *Person days*. *Vax per point*.

@@

### Summary of 2018 and 2019 campaigns
```{r}
species_2018 <- table(animals_2018$Species)
species_2019 <- table(tolower(animals_2019$`Dog or Cat`))
```

@@

*Table 1 should describe dogs!! (Take out cat parts & keep dogs only, & person days, split by location!)*

@@

### Comparing campaign costs and willingness to pay

The 2018 campaign cost more overall and per animal vaccinated than the 2019 campaign, largely due to increased personnel costs (Fig 3A & B, Table 1). Reflecting the extra personnel necessary to run the central point campaign, the 2018 campaign also took substantially more person-days per animal vaccinated (Table 1). We found that in most contexts, charging owners for vaccination would result in minimal cost recovery (Fig 3C), but that even a nominal fee would significantly reduce vaccination coverage acheived (Fig 3D), particulary in the rural commune of Andasibe.

```{r this is a code chunk!} 
knitr::include_graphics("figs/fig2.jpeg")
```

@@

In Andasibe Ville, we were only able to did this in one of our campaign locations, due to the sparsely distributed households and roads throughout the area, as such coverage in this. Make sure to mention back calc of pop & hdr is it reasonable?

```{r this is a code chunk!}
knitr::include_graphics("figs/fig1.jpeg")
```

### Dog demography and simulating vaccination strategies
For dogs, the sex ratio was male-biased (55%); while for cats it was female-biased (68% female). For dogs, the sex ratio was male-biased (65%), while for cats it was female-biased (66% female).

```{r this is a code chunk!}
knitr::include_graphics("figs/fig3.jpeg")
```

@@

Targeting dogs only?
Point to lit on willingness to pay / person day comp / cost per animal vaccinated

PrEP and exposures during vax, we had N exposures. Puppies were the easiest to handle for both owners and vaccinators and vax is effective in them. Better dosing guidelines. Longer lasting vax.

More community based delivery (Ahmed's paper, Pivot, accessibility, etc.). Possibly combining different campaign strategies could work per models Turnover is tricky & waning doesn't contribute too much. Thermotolerant & 3 yrs even without booster. (although most dogs are revaccinated and we do this in the model). Volunteer based/NGO based. Push for govt investment and viewing as a public good. Modeling after childhood vax = routine through local offices of puppies + supplemental activities by trained vets annually to catch unvax (feral, oral). The veterinary workforce (relieving pressures on them, challenges with private-public roles).
Paying people is important hard/relieving pressures and making sure it's viewed as a public good (Ahmed's paper). Veterinary students?

@@

Policy recommendations:
- Dog vax as a public good, remove barriers for dog owners & incentivize vets/etc/ remove economic pressures
- Community based vax delivery particulalry in novel areas (could lower cost while improving coverage)
- Easier ways to estimate coverage (& coupling with surveillance to assess impact)

@@

We thank vets livestock officers volunteers fokontany leaders government officials. We want to especially acknowledge the important contributions made by Annie Yang, an undergraduate researcher and volunteer with Princeton University and the Madagascar Dog Initiative who tragically passed away in a car accident before this research was published.
